Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-05-16 analysis/paper/paper.qmd
Updated affiliation

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials. 125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
---

@@

Highlights:

* Non-destructive analysis of prehistoric bifacial points made in quartzite using near-infrared spectroscopy, Raman scattering and X-Ray Fluourescence. 
* Each method detects differences in the structural and chemical characteristics in the material, but on their own the resolution is too low for predictive modelling. 
* Data fusion techniques are recommended for predictive modelling using non-destructive spectroscopy on quartzite material.

>>>>

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials. 125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
---

@@
